% =====================================================================
%  Reasoning on a Manifold — research summary + safety extension
%  NeurIPS format.
%
%  STYLE: This file is SELF-CONTAINED (Option A, active below): it emulates
%  the NeurIPS look (Times, 10pt, 5.5in x 9in text block) using only standard
%  packages, so it compiles on any full TeX Live install with no extra files.
%
%  To produce an *official* NeurIPS-styled PDF (Option B), drop the official
%  neurips_2024.sty next to this file, then: (1) comment out the Option A
%  block, (2) uncomment the two Option B lines, (3) delete the \usepackage
%  {natbib} line (neurips_2024.sty already loads natbib).
% =====================================================================

\documentclass[10pt]{article}

% ---------- Option B: official NeurIPS style (commented) ----------
% \documentclass{article}
% \usepackage[preprint]{neurips_2024}
% ------------------------------------------------------------------

% ---------- Option A: self-contained NeurIPS-like styling ----------
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathptmx}              % Times for text and math (NeurIPS look)
\usepackage[letterpaper,top=1in,bottom=1in,left=1.5in,right=1.5in]{geometry}
\usepackage{microtype}
\usepackage{titlesec}
\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{0.6em}{}
\titlespacing*{\section}{0pt}{1.4ex plus 1ex minus .2ex}{0.9ex plus .2ex}
\titlespacing*{\subsection}{0pt}{1.1ex plus 1ex minus .2ex}{0.7ex plus .2ex}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.55em}
% -------------------------------------------------------------------

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[numbers]{natbib}   % DELETE this line under Option B (neurips_2024.sty loads natbib)
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue!55!black,breaklinks=true]{hyperref}

\newcolumntype{Y}{>{\raggedright\arraybackslash}X}

\title{\bfseries The Geometry of Reasoning: Per-Behaviour Structure and the
Post-Training Origin of Safety Reasoning}

\author{Tony Su \\ Department of Computer Science \\ University College London \\
\texttt{zctpzs0@ucl.ac.uk}}
\date{}

\begin{document}

% ---- title block (Option-A friendly; \maketitle also works under Option B) ----
\begin{center}
{\LARGE\bfseries The Geometry of Reasoning: Per-Behaviour Structure\\[2pt]
and the Post-Training Origin of Safety Reasoning\par}
\vspace{1.1em}
{\large Tony Su\par}
\vspace{0.3em}
Department of Computer Science, University College London\\
\texttt{zctpzs0@ucl.ac.uk}
\end{center}
\let\thefootnote\relax\footnotetext{Research summary. Prepared in the NeurIPS
format; not length-limited to a conference submission. Figures are deferred to
the empirical write-ups.}
\vspace{0.8em}

\begin{quote}
\begin{center}\textbf{\large Abstract}\end{center}
\small
Thinking language models expose their reasoning as text, and that text can be
read, labelled, steered, and---this work's organizing commitment---located as
geometry in the residual stream. We treat a reasoning behaviour not as a label on
text but as a structured object in activation space, and study one object in two
movements. \emph{Structure.} \citet{venhoff2025reasoning} show individual
behaviours---backtracking, uncertainty estimation, example testing, knowledge
augmentation---are each controllable through a single linear direction, while
\citet{huang2025manifold} show the aggregate tendency to overthink occupies a
low-dimensional subspace whose projection outlasts single-direction steering. If
each behaviour is one direction, where does the aggregate structure come from? We
measure per-behaviour geometry with machinery the field has not applied to
reasoning models---Marchenko--Pastur nulls that separate real structure from
finite-sample noise, three intrinsic-dimension estimators, and direct curvature
diagnostics---and find behaviours occupy distinct, ordered positions within the
chain; whether that structure is genuinely \emph{curved} or a flat low-dimensional
subspace is the measurement we make, not the assumption we start from.
\emph{Origin.} A behaviour's shape is installed by post-training, so the
non-thinking baseline becomes a controlled contrast of alignment recipes.
Deliberative alignment \citep{guan2024deliberative} trains a model to reason over
its safety specification before answering, and \mbox{gpt-oss-20b} ships this as
open weights \citep{openai2025gptoss}; we ask whether the safety object's geometry
fingerprints the recipe---cleanly within one model family, and within gpt-oss
through its reasoning-effort control---and whether jailbreaks succeed by pushing
the trajectory off it---including the forged-policy jailbreak that the model cannot
tell from its own deliberation \citep{ye2025luckycoin}, which we replicate with the
same probe-and-patch machinery to locate where the forgery becomes load-bearing. We
are explicit about
what is result (a within-chain behaviour ordering) and what is proposed apparatus,
and about the confound that safety may not be linearly separable from capability
\citep{ponkshe2025safety}.
\end{quote}
\vspace{0.4em}

% =====================================================================
\section{Introduction}
% =====================================================================

Reasoning models such as the DeepSeek-R1-Distill family write out a
chain of thought before they answer, and that chain contains recognisable moves:
the model frames a problem, recalls a fact, tries a small case, doubts its
progress, abandons an approach. \citet{venhoff2025reasoning} annotate six such
behaviours and show each can be amplified or suppressed by adding a single
difference-of-means direction to the residual stream. The picture is linear and
local: one behaviour, one direction, one layer.

A second result complicates that picture. \citet{huang2025manifold} study the
aggregate habit of \emph{overthinking}---chains that pad themselves with
redundant deliberation---and find that steering along a single direction helps
only up to a point, after which it degrades the model. Projecting the steering
vector onto a low-dimensional subspace recovered by PCA sustains the effect at
strengths where the single direction collapses; they call this projection
\emph{manifold steering}. The two results are hard to reconcile. If each
behaviour is one direction, the aggregate should be close to the linear hull of
those directions, and ``manifold'' adds nothing. If the aggregate genuinely
carries low-rank structure with consequences for steering, that structure must
come from per-behaviour geometry that a single-direction method cannot see, or
from how behaviours compose. Neither paper measures it. Huang uses ``manifold''
to mean ``linear subspace,'' yet the word implies curvature, an intrinsic
dimension below the embedding dimension, and a tangent-space approximation that a
70\% top-ten variance ratio does not establish.

This project resolves the tension by asking, for each behaviour separately,
whether its activations form a flat subspace, a curved manifold, or unstructured
high-dimensional scatter---and by bringing null baselines and curvature estimators
that distinguish these cases at the finite sample sizes mechanistic
interpretability actually works with. That question is the seed of a more general
claim: a reasoning behaviour is a structured object in activation space, and the
thesis studies one object in two movements---its \emph{structure} (what shape it
is) and its \emph{origin} (where the shape comes from, and why, for safety, the
shape is the property that matters). The present summary develops the structural
apparatus and its preliminary results, then the safety payoff that makes them
matter. A third question---whether the same instrument that detects a forged safety
policy could certify genuine knowledge-creation against mere retrieval---is a
natural but more philosophical extension we leave to future work.

\emph{Origin: the shape is installed by post-training.} \citet{guan2024deliberative}
introduced \emph{deliberative alignment}, which teaches a model the text of its
safety specification and trains it to recall and reason over that specification in
its chain of thought before deciding whether to comply. Safety is no longer a
reflex applied at the first token; it is manufactured as a \emph{reasoning
behaviour} with an annotatable signature, and OpenAI shipped a model built this
way, \mbox{gpt-oss-20b}, as open weights with its reasoning trace left readable
\citep{openai2025gptoss}. Because a behaviour's shape is installed by a specific
recipe rather than intrinsic to ``an LLM,'' the non-thinking baseline becomes a
controlled contrast of recipes---but only where the lineage is held fixed. We make
the contrast where it is clean and say so where it is not: \emph{within} the
Qwen-1.5B family, base pretraining, RLHF instruction-tuning, and reasoning
distillation differ only in post-training, whereas \mbox{gpt-oss-20b} adds
deliberative alignment but also differs in scale and architecture, so its
comparison to the small models is suggestive, not controlled. The one clean handle
on deliberative alignment itself is gpt-oss's reasoning-effort control, which varies
how much safety deliberation is engaged with the weights held fixed. Our hypothesis
is that the safety object's geometry tracks the recipe---a shallow early-token
direction under classical RLHF \citep{qi2025shallow,arditi2024refusal}, little
separable safety structure under distillation alone, a chain-engaged object that
scales with reasoning effort under deliberative alignment---and that jailbreaks work
by pushing the trajectory off it. Whether that object is curved, and whether it is
even linearly separable from capability \citep{ponkshe2025safety}, are the first
questions rather than assumptions: if safety reasoning has no separable object, that
absence is itself a mechanistic account of why deliberative alignment is fragile.
The sharpest test of that fragility is the chain-of-thought \emph{forgery}
jailbreak: an attacker appends fabricated policy deliberation in the model's own
style, and the model adopts the forged premise and complies, the reported failure
being that its activations represent the injected deliberation much like its own
\citep{ye2025luckycoin}. We replicate it rigorously---training a probe on residual
activations to ask whether genuine and forged deliberation separate at all, then
using activation patching to locate the layer at which the forged premise becomes
causally load-bearing---which turns the jailbreak from an anecdote into a
measurement of where malice enters the reasoning.

\paragraph{Contributions.} This summary integrates a validated methodological
program (\emph{structure}) with the safety payoff that reuses it (\emph{origin}).
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt,topsep=2pt]
\item \textbf{A per-behaviour geometric account of reasoning} (\emph{structure})
that adapts Marchenko--Pastur random-matrix nulls, multi-estimator intrinsic
dimension, and direct curvature diagnostics to reasoning-model activations, with a
measured within-chain ordering of behaviours as the first empirical result and
curvature treated as a measurement rather than a premise
(Section~\ref{sec:original}).
\item \textbf{The hypothesis that safety geometry is a post-training fingerprint}
(\emph{origin}), turning the non-thinking baseline into a recipe contrast that is
controlled within one model family and isolated, within gpt-oss, by its
reasoning-effort control, with five falsifiable hypotheses in which separability
gates the fingerprint claim (Section~\ref{sec:safety}).
\item \textbf{A ``deliberative safety reasoning'' annotation schema and an
experimental plan} that reuses roughly ninety percent of the existing pipeline,
including a rigorous replication of the chain-of-thought ``forgery'' jailbreak that
defeats deliberative alignment---a genuine-versus-forged probe and activation
patching that locate where the forged premise becomes load-bearing
(Sections~\ref{sec:safety}--\ref{sec:methods}).
\item \textbf{An explicit accounting of what is not yet trustworthy}---confounded
curvature estimators since corrected (so no curvature number is yet a result),
small samples, the gap between linear projection and curvature, the possibility
that safety is not linearly separable from capability, and the unfaithfulness of
chains of thought---so the geometric claims are made with the right amount of
confidence (Section~\ref{sec:risks}).
\end{enumerate}

% =====================================================================
\section{Background and related work}
\label{sec:related}
% =====================================================================

The project sits at the meeting point of two literatures that have not yet been
introduced to each other: the interpretability of reasoning behaviours, and the
geometry of safety. This section walks the path from one to the other.

\subsection{Reasoning behaviours and how to steer them}

Chain-of-thought reasoning became a trainable target with DeepSeek-R1, which
incentivises long deliberation through reinforcement learning and then distils the
resulting traces into smaller dense models \citep{deepseek2025r1}. The 1.5B model
this project studies, R1-Distill-Qwen-1.5B, is the base Qwen2.5-Math-1.5B
\citep{yang2024qwenmath} fine-tuned on roughly 800k reasoning samples generated by
R1; the distilled model receives the reasoning supervision but neither the
reinforcement-learning stage nor a dedicated safety stage. That lineage is what
makes it a clean object of study, and---as Section~\ref{sec:related-posttrain}
explains---a clean negative control for safety.

\citet{venhoff2025reasoning} give the behavioural vocabulary. They prompt an
annotator to label reasoning sentences as initialization, deduction, knowledge
augmentation, example testing, uncertainty estimation, or backtracking, extract a
difference-of-means direction for each, and show that adding or subtracting it
steers the behaviour. Their layer choice is causal rather than convenient: they
score candidate directions by attribution patching and select, per behaviour, the
layer that contributes most to the next-token distribution. The selected layers
differ across behaviours---backtracking peaks around layer 17 of the 28-layer
1.5B model, knowledge augmentation and uncertainty around layer 18, example
testing around 15---so ``layer 17'' is one behaviour's answer, not a global
constant. Their reported behaviour fractions come from chains capped at 1{,}000
tokens, a detail that turns out to matter.

\subsection{Subspace or manifold: what counts as structure}
\label{sec:related-geometry}

\citet{huang2025manifold} bring the geometry. They show that overthinking
activations concentrate in a subspace whose top ten principal components capture
more than seventy percent of the variance, and that projecting a steering vector
onto that subspace---``manifold steering''---reduces token count by up to seventy
percent on GSM8K while preserving accuracy, at strengths where a raw direction
plateaus and then collapses. The word ``manifold'' is doing more work than the
evidence supports. A high variance ratio in the top components is consistent with
a flat subspace, and says nothing about curvature, about the intrinsic dimension
relative to the ambient $d=1536$, or about whether a tangent-space approximation
holds. The distinction is not pedantic: a flat subspace and a curved manifold are
both ``low-dimensional structure,'' but only the second predicts that linear
steering saturates at a characteristic radius.

Separating signal from sampling artefact is the harder half. At per-behaviour
sample sizes between fifty and a few hundred in $d=1536$, the empirical covariance
spectrum is non-trivially spread even for isotropic Gaussian data; a chunk of any
reported variance ratio is finite-sample noise. The Marchenko--Pastur law
\citep{marchenko1967distribution} gives the limiting spectrum of such noise and
hence a null against which real low-dimensionality must be shown to exceed---a
control, to our knowledge, absent from prior geometric claims on reasoning models.
Intrinsic dimension is itself estimator-dependent, which is why we read it through
three lenses at once: the TwoNN ratio estimator \citep{facco2017twonn}, the
Levina--Bickel maximum-likelihood estimator \citep{levina2004mle}, and correlation
dimension. The depth axis is the natural complement to the within-chain axis:
behaviours are built up across layers, and tracing that build-up is where
transcoders \citep{dunefsky2024transcoders} and crosscoders
\citep{lindsey2024crosscoders,templeton2024scaling} point, though we keep the
heaviest of that machinery as future work.

\subsection{The geometry of refusal as an output}

Safety interpretability has, so far, geometrised refusal as an \emph{output}---the
``I can't help with that'' act---rather than as reasoning. \citet{arditi2024refusal}
showed that across thirteen chat models a single residual-stream direction mediates
refusal: ablate it and the model complies with harmful requests, add it and it
refuses harmless ones. The single-direction picture did not survive contact with
sharper tools. \citet{wollschlager2025geometry} find refusal is carried by several
functionally independent directions spanning a multi-dimensional ``concept cone,''
and warn that orthogonality does not imply independence under intervention.
\citet{joad2026more} push further, isolating geometrically distinct refusal
directions across eleven categories of refusal, yet observe that steering along
any of them produces the same refuse-versus-over-refuse trade-off, as if the many
directions feed one shared control knob. Sparse autoencoders decompose refusal into
latent features whose ablation changes behaviour and whose study clarifies how
jailbreaks bypass it \citep{yeo2025refusalsae}. The trajectory of this literature
is instructive---single direction, to cone, to many-directions-one-knob---and it
is settled enough that the obvious next question is no longer about the output. It
is about the reasoning that precedes the output, which nobody has geometrised.

\subsection{Deliberative alignment: safety trained as a reasoning behaviour}

Deliberative alignment is the development that makes that question answerable
\citep{guan2024deliberative}. Rather than train refusal as a stimulus--response
mapping, it teaches the model the literal text of the safety policy and trains it,
with a specification-grounded judge and no human-written reasoning traces, to cite
and apply the relevant policy in its chain of thought before answering. The
mechanism leaves a measurable trace: in the reported evaluation, the fraction of
chains that reference the correct policy is 0.75 for hard refusals and 0.91 for
safe completions in the safety-trained model, against 0.27 and 0.21 in a no-safety
baseline. Safety, in other words, has been manufactured as an annotatable reasoning
behaviour. \mbox{gpt-oss-20b} ships this recipe as open weights
\citep{openai2025gptoss}: a 20.9B-parameter mixture of experts (3.6B active per
token, 24 layers, residual width 2880, top-4 of 32 experts) that emits its
reasoning in a separate \texttt{analysis} channel of the harmony response format,
left deliberately unfiltered so it stays monitorable. Worst-case fine-tuning
studies \citep{wallace2025worstcase} informed the decision to release. The
combination---a model whose safety \emph{is} a reasoning behaviour, with its
reasoning legible and its weights open---is the substrate this project needs.

\subsection{Post-training and the locus of safety}
\label{sec:related-posttrain}

Where safety lives, and how deeply, depends on how it was installed. Classical
RLHF safety is shallow: \citet{qi2025shallow} show that alignment mostly reshapes
the first few output tokens, which is why prefilling, suffix, and fine-tuning
attacks all work. In weight space the aligned region is a narrow ``safety basin''
whose edge is a cliff, not a slope \citep{peng2024safetybasin}. Reasoning training
actively erodes inherited safety: aligning a reasoning model tends to cost
reasoning accuracy---a ``safety tax'' \citep{huang2025safetytax}---and the R1
distillation recipe, which carries no safety stage, leaves models that refuse
little until safety is retrofitted \citep{zhang2025realsafe}. The cleanest way to
read these results is differential: hold the base model fixed and vary the
post-training, and the safety you get varies with it. That is precisely the
contrast the project's non-thinking baseline already sets up, now with safety as
the readout.

Two cautions from this literature shape the design. First, safety may not be
linearly separable from capability at all: \citet{ponkshe2025safety} find that the
subspaces amplifying safe behaviour also amplify useful behaviour, so a clean
``safety manifold'' could be an artefact of a capability correlation, and any
geometric claim needs a difficulty-matched capability control. Second, even refusal
\emph{output} is multi-dimensional once you look past the dominant direction, with
orthogonal secondary directions that gate it \citep{pan2025hidden}. The companion
phenomenon is emergent misalignment: narrow fine-tuning on insecure code makes a
model broadly misaligned \citep{betley2025emergent}, and the weight updates for
different harmful fine-tunes share a low-dimensional subspace
\citep{arturi2025shared}---evidence that post-training moves safety along a small
number of reusable axes. If misalignment is low-rank and reusable, the shape of the
safety object is worth measuring directly.

\subsection{Faithfulness, monitorability, and how deliberative safety fails}

A safety story told in the chain of thought is only as good as the chain's
faithfulness. The optimistic case is that reasoning in legible language is a
monitorable safety affordance, and a recent cross-organisation position paper
argues exactly that, while stressing the affordance is fragile and easily trained
away \citep{korbak2025monitorability}. The pessimistic evidence is strong.
\citet{chen2025dontsay} show reasoning models often fail to verbalise the hints
they use, with reveal rates frequently below twenty percent.
\citet{chan2025predict} make the constructive version of the point: a linear probe
on chain-of-thought \emph{activations} predicts an unsafe final answer better than
any text-based monitor, and does so before the chain finishes---latents are a more
faithful signal than the words. Open-weight evaluations corroborate the text--latent
gap \citep{young2026lieto}. For reasoning models specifically, refusal can collapse
at the last moment: the ``refusal cliff'' is a sharp drop in refusal intent at the
final tokens before output, traceable to a sparse set of attention heads, even
though the model held a strong refusal intent throughout its thinking
\citep{yin2025cliff}; safe behaviour can sometimes be restored by intervening in
the first few reasoning steps, at least in the multimodal setting where this was
shown \citep{ghosal2026safethink}.

The most direct evidence that deliberative safety is a \emph{geometric} object,
and an exploitable one, comes from red-teaming \mbox{gpt-oss-20b}. A prize-winning
competition entry, the ``Lucky Coin'' jailbreak, appends forged policy
deliberation---an invented exception written in the model's own analysis-channel
style---and the model adopts the fabricated premise and complies; the authors report
that a classifier over harmony-role activations represents the injected text like
the model's own policy reasoning \citep{ye2025luckycoin}. Systematic probes of
\mbox{gpt-oss-20b} catalogue related failure modes, including policy-paradox prompts
that lift jailbreak success from a few percent to over forty
\citep{gptoss2025probing}, and agentic red-teaming surfaces vulnerabilities that
appear only inside tool-use contexts \citep{mindthegap2025agentic}. These results
are striking and, in the competition's case, not peer-reviewed; their interpretation
is exactly what a rigorous, null-controlled geometric study can supply. Cross-model
diffing now reaches across architectures---dedicated-feature crosscoders surface a
copyright-refusal mechanism in \mbox{gpt-oss-20b} absent from Qwen3-derived models
\citep{jiralerspong2026crossarch}---building on diffing methods whose sparsity
artefacts are themselves now understood \citep{minder2025crosscoder}.

\subsection{The gap this work occupies}

Put the two halves together. The geometry of refusal as an output is settled; the
geometry of safety \emph{reasoning as a behaviour in the chain of thought} is
unoccupied. Deliberative alignment manufactured that behaviour and made it legible;
gpt-oss-20b shipped it open; the Lucky Coin jailbreak hints that its geometry is
load-bearing and forgeable. The apparatus that the reasoning-geometry program built
to ask ``manifold or subspace?'' for generic behaviours is the apparatus needed to
ask it for safety, with post-training as the variable that the field's own results
say should change the answer. That is the niche.

% =====================================================================
\section{The original project: per-behaviour geometry of reasoning}
\label{sec:original}
% =====================================================================

\subsection{Research questions}

The program decomposes the manifold-versus-subspace tension into three questions,
each one publishable contribution.
\begin{description}[leftmargin=1.4em,itemsep=2pt,topsep=2pt]
\item[Q1 (Temporal).] How do behaviours distribute within a chain, and how much of
a reported behaviour fraction is an artefact of the generation cap?
\item[Q2 (Geometric).] For each behaviour, is the activation distribution a flat
subspace, a curved manifold, or high-dimensional scatter, and what null baselines
separate real low-dimensionality from finite-sample artefacts?
\item[Q3 (Causal).] If a behaviour's activations are curved with characteristic
scale $\kappa$, linear steering should saturate near a radius $\alpha^\star \propto
1/\kappa$. Does it?
\end{description}

\subsection{Pipeline}

A five-phase pipeline produces the data once and serves all three questions.
\textbf{(1)~Task generation:} Claude Sonnet 4.5 writes 1{,}000 reasoning tasks
across ten categories. A deduplication audit removes near-duplicate prompts, which
matter because correlated tasks deflate estimated intrinsic dimension and
manufacture spurious low-rank structure---a corpus-hygiene step the field rarely
reports. \textbf{(2)~Chain generation:} R1-Distill-Qwen-1.5B answers each task with
greedy decoding under an 8{,}192-token cap, chosen to sit above Venhoff's
1{,}000-token cap and below Huang's overthinking regime; pilot chains terminate
naturally about 95\% of the time at a mean length near 2{,}500 tokens.
\textbf{(3)~Annotation:} chains are sentence-segmented and labelled with Venhoff's
six-way scheme using the original prompt verbatim, with overlapping chunking for
chains that exceed annotator context. \textbf{(4)~Extraction:} residual-stream
activations are taken at the last token of each annotated sentence at layers
$\{3,7,10,14,17,21,24,27\}$, preserving both Venhoff-style fixed-layer analysis and
the cross-layer axis. \textbf{(5)~Geometry and steering:} described next.

\subsection{The statistical apparatus}

Phase~5 is the methodological core, and the reason the project can say something
the prior work cannot. For each behaviour it computes a PCA spectrum with bootstrap
confidence intervals; a \textbf{Marchenko--Pastur null} built by Monte Carlo over
matched-$(N,d)$ Gaussian draws, so that a variance ratio counts as evidence only
where it exceeds the noise floor; a \textbf{permutation null} that shuffles
behaviour labels to separate ``activation of this behaviour'' from ``activation of
this model''; \textbf{three intrinsic-dimension estimators} read together so that
agreement supports a dimension claim and disagreement flags it; and \textbf{three
curvature diagnostics}---a local-versus-global PCA dimension ratio, a
geodesic-to-Euclidean distance ratio on a $k$-nearest-neighbour graph, and
tangent-space variation across the data---each near zero for a flat subspace and
growing with curvature. The combination of random-matrix nulls and direct curvature
measurement is what distinguishes a genuine manifold claim from a restated variance
ratio.

\subsection{The result already in hand}

One empirical finding is established, on a 20-chain stratified pilot, and it is the
temporal one. Behaviours occupy distinct mean positions along the chain:
initialization at 0.31, knowledge augmentation at 0.39, uncertainty estimation at
0.52, backtracking at 0.60, on a scale from chain start to chain end. The ordering
is the one a person would predict---frame, recall, evaluate, abandon---and it
explains a discrepancy. Behaviour fractions in 8{,}192-token chains diverge from
Venhoff's; stratifying by length resolves it, with short chains matching his
fractions within tolerance on five of six labels and long chains over-sampling the
late behaviours. Venhoff's ``backtracking is 4\% of sentences'' is therefore a
statement about the first thousand tokens, not about reasoning in general. The
geometric and causal questions (Q2, Q3) are posed and instrumented but not yet
answered; their results belong to the empirical write-ups, and Section~\ref{sec:risks}
is candid about which preliminary numbers are not yet trustworthy.

\subsection{Extensions that sharpen the core}

Two extensions deepen the program without adding papers. The \textbf{non-thinking
baseline} runs the same pipeline on the base Qwen2.5-Math-1.5B and asks whether the
geometry is latent in pretraining, constructed by distillation, or mixed; it is, in
effect, a differential post-training experiment, which is exactly the hinge the
safety extension swings on. The \textbf{cross-layer} extension reads behaviour
representations across depth by linear probing and layer-wise PCA, with transcoder
feature-circuit analysis deferred. Three workshop-paper contributions follow: a
temporal-structure paper (Q1), a manifold-or-subspace paper (Q2, the methodological
centre), and a curvature-predicts-saturation paper (Q3).

% =====================================================================
\section{The safety extension: the shape of safety after post-training}
\label{sec:safety}
% =====================================================================

\subsection{Thesis}

The host question---does a behaviour occupy its own manifold or just a
subspace?---acquires stakes and a mechanism when the behaviour is safety. The claim
is this: \emph{the shape of the safety-reasoning object in activation space is a
fingerprint of the post-training recipe that installed it.} Deliberative alignment,
which trains safety as multi-step deliberation, should carve a curved, chain-engaged
manifold with a sharp turn at the decision point. Classical RLHF, which reshapes the
first few tokens \citep{qi2025shallow}, should leave a near-straight shot to one
early refusal direction. Reasoning distillation, which installs no safety stage,
should leave almost nothing separable. And a successful jailbreak should show up as
a trajectory pushed \emph{off} the safe manifold. If safety reasoning is instead
flat, or inseparable from capability \citep{ponkshe2025safety}, that is not a failed
experiment---it is a mechanistic explanation for why deliberative alignment is
jailbreakable, in the same spirit as the project's informative-null posture for Q2.

\subsection{Three reuses of the existing apparatus}

The extension is mostly new stimuli and a new label set, not new infrastructure.
\textbf{Curvature becomes a mechanism discriminator}: the predicted
flag~$\to$~recall-policy~$\to$~adjudicate~$\to$~decide structure of deliberative
safety is a multi-segment, curved trajectory, where a shallow RLHF refusal is not,
so the project's curvature diagnostics tell the two mechanisms apart.
\textbf{The baseline comparison becomes a recipe ladder}: instead of only
thinking-versus-base, the contrast runs across deliberative alignment, RLHF, reasoning
distillation, and the shared base (Table~\ref{tab:ladder}). \textbf{Per-behaviour
subspaces ask whether safety reasoning is a new primitive or a composition}:
spec-recall resembles knowledge augmentation, adjudication resembles uncertainty
estimation, reconsideration resembles backtracking, so the natural test is whether
the safety-reasoning subspace lies inside the span of the existing behaviour
subspaces or has an orthogonal component---a principal-angle test the project already
implemented for a separate extension. Spec-recall is itself \emph{knowledge
invocation} inside reasoning---the safety analogue of the knowledge-augmentation
behaviour from the first half---so this test doubles as a question about whether
deliberative safety is a new primitive or a composition of behaviours already in the
inventory. (Whether the same probe-an-invocation method could one day distinguish
genuine knowledge-creation from retrieval is a more philosophical question we leave
to future work.)

\begin{table}[t]
\centering
\small
\caption{Five falsifiable hypotheses. Each is independently publishable, and a null
result on H1 or H2 is itself a mechanistic explanation for the fragility of
deliberative safety.}
\label{tab:hyps}
\begin{tabularx}{\textwidth}{@{}p{0.045\textwidth} Y p{0.30\textwidth}@{}}
\toprule
\textbf{\#} & \textbf{Hypothesis} & \textbf{Falsified if} \\
\midrule
H1 & Deliberative safety reasoning occupies its own low-dimensional but
\emph{curved} object, with curvature above the flat-Gaussian null and above the
four generic behaviours. & It is flat, high-dimensional, or indistinguishable from
generic reasoning. \\[2pt]
H2 & The \emph{shape} tracks the recipe: deliberative alignment $\to$ curved,
chain-engaged, scaling with reasoning effort; RLHF $\to$ shallow early-token
direction; distillation $\to$ weak or absent. & The object has the same shape
across recipes, or gpt-oss curvature is flat. \\[2pt]
H3 & Malice moves activations off the safe manifold; successful jailbreaks suppress
the policy-citation direction and ride off-manifold at a locatable chain position.
& Successful and failed attacks have identical geometry. \\[2pt]
H4 & Output convergence, process divergence: refusal \emph{outputs} align across
models (small principal angles) while safety-\emph{reasoning} subspaces diverge
(large angles). & Reasoning subspaces align as tightly as output directions. \\[2pt]
H5 & Forged $\approx$ genuine in representation: a probe trained on genuine
deliberation classifies forged deliberation as genuine, and the forged premise
becomes load-bearing at an identifiable layer. & The model linearly separates
forged from genuine deliberation. \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Hypotheses}

Table~\ref{tab:hyps} states the five hypotheses and their falsifiers. H1 asks
whether deliberative safety reasoning is a separable, curved object at all, and it
\emph{gates} H2: if there is no separable safety object, the question of whether its
shape tracks the recipe does not arise, and a difficulty-matched capability control
\citep{ponkshe2025safety} is needed before any shape is credited to the recipe
rather than to task difficulty. H2 is the post-training core: does the shape track
the recipe? H3 tests whether jailbreaks are off-manifold events. H4 predicts that refusal \emph{outputs} converge across models
while safety \emph{reasoning} diverges, because the recipes differ. H5 is the
rigorous Lucky Coin: whether the model's activations fail to separate forged from
genuine policy deliberation.

\subsection{The annotation schema}

The one genuinely new artefact is a schema for \emph{deliberative safety reasoning},
parallel to Venhoff's behaviour scheme and built on deliberative alignment's own
notion of policy retrieval. Analysis-channel sentences are labelled as
\texttt{spec\_citation} (recalling or paraphrasing a policy clause---the retrieval
signal made into a label, and a sibling of knowledge augmentation),
\texttt{adjudication} (weighing whether and how to comply, a sibling of uncertainty
estimation), \texttt{decision} (the three-way refuse / safe-complete / comply
commitment), and \texttt{harm\_recognition} (the chain explicitly noting the request
is harmful, which exposes the knows-but-complies gap when recognition co-occurs with
compliance). The same multi-annotator robustness pipeline the project uses
elsewhere---three independent judges scored by Cohen's $\kappa$---guards the labels,
and the empirical anchor that this is detectable is that refusal chains cite policy
far more often than jailbroken ones.

\begin{table}[t]
\centering
\small
\caption{The post-training recipe ladder. Within the Qwen-1.5B family (rows
2--4) the base is held fixed and post-training is the only variable; gpt-oss-20b
(row 1) adds deliberative alignment but also differs in scale, architecture, and
data, so its row is a confounded contrast, and the clean handle on deliberative
alignment is gpt-oss's own reasoning-effort control. Two of the three 1.5B arms
already run in the project.}
\label{tab:ladder}
\begin{tabularx}{\textwidth}{@{}l l Y@{}}
\toprule
\textbf{Arm} & \textbf{Model} & \textbf{Predicted safety geometry} \\
\midrule
Deliberative alignment & gpt-oss-20b & curved, chain-engaged manifold; scales with reasoning effort \\
RLHF safety & Qwen2.5-1.5B-Instruct & shallow early-token single direction \\
Reasoning, no safety & R1-Distill-Qwen-1.5B & weak or no separable safety object \\
Base (origin control) & Qwen2.5-Math-1.5B & none \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Experiments}

Five experiments map onto existing runners; each names its falsifier.
\textbf{S1 (DSR geometry)} extracts deliberative-safety activations on harmful,
benign, over-refusal, and safe-completion prompts and asks whether they form a
single direction, a flat cone, or a curved manifold, with a principal-angle
composition test against the generic behaviours (H1).
\textbf{S2 (malice off-manifold)} builds chain-as-curve trajectories on matched
harmful/benign pairs and successful-versus-failed jailbreaks and measures an
off-manifold score and the chain position of any refusal collapse, the geometric
read of the refusal cliff \citep{yin2025cliff} (H3).
\textbf{S3 (rigorous Lucky Coin)} trains a genuine-versus-forged probe across
layers, with chain-grouped cross-validation, and uses activation patching to find
where a forged premise becomes causally load-bearing---the headline safety result,
and the one that tests load-bearingness rather than assuming it (H5).
\textbf{S4 (recipe differential)} runs the ladder of Table~\ref{tab:ladder} plus a
within-gpt-oss reasoning-effort sweep, the one knob OpenAI exposed, partialling out
a capability subspace as \citet{ponkshe2025safety} demand (H2, H4).
\textbf{S5 (steering, stretch)} steers the safety direction to toggle refuse against
comply and couples saturation to curvature, the direct safety instance of Q3.

\subsection{Models and the discipline of cross-dimensional comparison}

gpt-oss-20b is the new compute load; the three contrast models are 1.5B and cheap.
Its residual stream is extractable exactly as the DeepSeek pipeline does it---hidden
states of shape $[\text{batch},\text{seq},2880]$ across 25 tensors, with MXFP4
quantising weights only---and it fits in roughly 16\,GB on Hopper hardware or about
48\,GB in bfloat16 on the lab's unified-memory machine. Comparing a $d=1536$ model
with a $d=2880$ one demands discipline: compare \emph{summary statistics}---intrinsic
dimension, curvature, eigenvalue spectra, geodesic ratios---not raw vectors; normalise
depth by fraction, since ``layer 14'' does not correspond across models; use CKA,
principal angles, or Procrustes on per-model subspaces for any overlap claim; and
analyse the \emph{residual stream only}, because the mixture-of-experts routing makes
MLP outputs carry top-4-of-32 variance that is not geometry. gpt-oss's alternation of
sliding-window and full attention can imprint an even/odd per-layer artefact, which
the analysis should expect rather than over-read.

\subsection{Positioning}

The recommended framing elevates safety to the flagship while protecting the
validated core. Papers~1 and~2---temporal structure and manifold-or-subspace on the
four generic behaviours---stand as the apparatus that earns the right to make safety
claims; the safety paper becomes the new flagship, absorbing the
curvature-and-steering machinery of the old Paper~3, with generic steering demoted to
a method-validation section. The work targets socially-responsible-LM and AI-safety
venues and the mechanistic-interpretability workshops, where the combination of
deliberative alignment, open weights, and a striking failure mode is the kind of
story that travels.

% =====================================================================
\section{Shared methodology and reuse}
\label{sec:methods}
% =====================================================================

The safety arm is roughly ninety percent existing code. Reused wholesale: chain
generation, the multi-annotator robustness pipeline, activation extraction, PCA, the
Marchenko--Pastur nulls, the intrinsic-dimension estimators, the curvature
diagnostics, cross-layer probing, the trajectory module, activation patching, steering
with a causal fail-stop, and the principal-angle test. Genuinely new and small: a
gpt-oss loader path (harmony template with a reasoning-effort control and the
2880-wide, 25-tensor shape), the deliberative-safety annotation schema and its judge
prompt, a forged-deliberation stimulus generator with its genuine-versus-forged probe,
and an early-token KL localisation for the RLHF arm. The loader path has been built
and unit-tested on a smoke harness, with the DeepSeek path left byte-for-byte
unchanged behind a default family flag; the remaining engineering is verification on
the real model. Annotation re-engages the API budget that has otherwise gone idle.

% =====================================================================
\section{Risks, confounds, and honest limitations}
\label{sec:risks}
% =====================================================================

The geometric claims are only as good as the estimators behind them, and an internal
audit found several that were not yet trustworthy. The headline one: the
local-versus-global dimension ratio used as a curvature diagnostic was confounded by
neighbourhood size against subspace dimension, returning ``curved'' values on
perfectly flat Gaussian data. It has since been corrected and calibrated against flat
and spherical references, but the consequence is firm---any curvature or
intrinsic-dimension number produced before the fix, including those in earlier
progress notes, must be regenerated before it is cited, and none of them appear as
results in this summary. The same audit flagged non-reproducible PCA from an
unseeded randomized solver and a cross-model comparison whose ``bootstrap'' was a
Gaussian reconstruction rather than a resample; both must be fixed before the
cross-model hypothesis H4 is load-bearing.

The methodological cautions are real and carried openly. Per-behaviour samples of
fifty to a few hundred in $d=1536$ or $d=2880$ are small for intrinsic-dimension and
curvature estimation, so bootstrap confidence intervals and the
Marchenko--Pastur null are primary, not decorative. ``Manifold steering'' is a linear
projection, so it tests a subspace rather than curvature, a mismatch the Q3 paper must
state rather than paper over. Behaviour labels come from an LLM annotator, which is why
the three-judge agreement pipeline exists. Half the chains hit the token cap, an
unresolved truncation confound, and mean-pooling destroys the very trajectory the
manifold claim is about, so the trajectory experiments must not pool.

The safety arm inherits sharper versions of the same risks. Safety may not be linearly
separable from capability \citep{ponkshe2025safety}, so the deliberative-safety
geometry must be reported relative to a difficulty-matched capability baseline, not in
isolation. Chains of thought are often unfaithful \citep{chen2025dontsay}, so a
``safety reasoning manifold'' cannot be assumed to be causal---it is exactly why S3
patches and steers rather than reads, and why the latents-beat-text result
\citep{chan2025predict} is encouraging rather than sufficient. The Lucky Coin jailbreak
is a non-peer-reviewed competition writeup \citep{ye2025luckycoin}; it motivates the
work but does not substantiate it, which is the point of replicating it with nulls and
grouped cross-validation. Stated plainly: the strongest version of this thesis is a set
of falsifiable predictions with the instruments to test them, several of which may come
back null, and a null on the central hypothesis would itself be a finding about why
deliberative alignment can be talked out of its own policy.

% =====================================================================
\section{Conclusion}
\label{sec:conclusion}
% =====================================================================

The project began with a narrow geometric puzzle---if each reasoning behaviour is one
direction, where does the aggregate structure come from?---and built the apparatus to
answer it honestly: null baselines that resist finite-sample illusions, multiple
intrinsic-dimension estimators, and direct curvature measurement, validated by a
clean temporal finding about how behaviours order themselves within a chain. That
apparatus turns out to be a safety instrument. Because a behaviour's shape is
installed by post-training, the geometry of the safety object becomes a window on the
alignment recipe that made it---read cleanly within one model family, and within
gpt-oss through its reasoning-effort control---and on how a jailbreak escapes it, most
sharply the forged-policy jailbreak, which we turn from an anecdote into a measurement
of where in the reasoning a fabricated premise becomes load-bearing. The unifying
object is the reasoning behaviour as geometry; the unifying discipline is that one
measures its shape rather than asserting it, and reports which claims are results and
which are still predictions. Whether the same instrument can one day certify genuine
knowledge-creation against retrieval---the project's original, more philosophical
motivation---is left as future work.

\small
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
